\thetitle

### 2021-03-04 writing/challenges.tex
issues with sections 9, 10, and 11, plus triggers for global searches

While at first glance the method of section~\ref{sec:why-rational} seems
straightforward to implement and test, three challenges remain. The first is
comparing the effect of blame on the rational programmer across three different
mechanisms; second is the collection of a large number of representative
debugging scenarios; and third is finding a proper sampling strategy for the
huge space of possibilities. A coincidental challenge is the disparity of the
implementations of gradually typed languages. To eliminate this variable, the
authors use Racket, which is thus far the only language in which all three major
soundness mechanisms are available in a robust and comparable manner; for the
implementation of Transient Typed Racket, see the parallel submission of an
experience report~\cite{ttt21}.

>>>>

While at first glance the method of section~\ref{sec:why-rational} seems
straightforward to implement and test, three challenges remain. The first is
comparing the effect of blame on the rational programmer across three different
mechanisms; second is the collection of a large number of representative
debugging scenarios; and third is finding a proper sampling strategy for the
huge space of possibilities. A coincidental challenge is the disparity of the
implementations of gradually typed languages. To eliminate this variable, the
authors use Racket, which is thus far the only language in which all three major
soundness mechanisms are available in a robust and comparable manner:
Typed Racket implements Natural, Shallow Racket~\cite{ttt21}
Transient, and plain Racket for Erasure.

### 2021-03-04 writing/conclusion.tex
touch ups for consistent use of mutant/mutated module and module/component
issues with sections 9, 10, and 11, plus triggers for global searches

The paper provides the first insight into the measurable value of blame in the
gradually typed world. In an implicit manner researchers and language designers
have answered this question one way or another, but without evidence. Now they
may use the evaluation method of this paper to rationalize their answers.  The
work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers prefer
the run-time checking of Natural over other soundness methods. But, just because
a random set of programmers expresses this view does not mean that blame
assignment adds value. At this point, it is clear that, in some circumstances,
it does and in others, the answer remains murky.

The paper does {\em not\/} address a problem in the gradually typed world that
was pointed out early on by practical researchers~\cite{incorrect-ts,
sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
attention~\cite{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
themselves.  Developers use gradual typing to move an untyped code base into the
typed realm, and to this end, they need typed APIs for the vast repositories of
already-existing libraries. Because they can, language implementors merely
create typed facade modules that import untyped functions and export them with
type annotations---instead of converting the libraries themselves. With those
facades, the compiler can easily type-check typed components. The problem is
that these retroactive additions of types to a library may result from a
misunderstanding of the code. \emph{The ascribed types may thus be a mistake
themselves.}

The cited evidence suggests that this scenario is quite common and largely
unadressed.  A future evaluation must develop mutators that produce incorrect
type annotations without breaking the code itself. Some preliminary work
suggests that such type mutators are even more difficult to develop than the
type-level code mutators presented here. Based on the gradual typing literature,
the Natural semantics should usually discover such mistakes. In contrast, the
Transient and Erasure semantics cannot help with such mistakes at all; indeed,
we expect that these latter two raise misleading exceptions or produce
plainly-wrong answers.  Of course, only further evaluation work can confirm or
reject these conjectures.

>>>>

The work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers
prefer the run-time checking of Natural over other soundness methods. But just
because a random set of programmers expresses this view does not mean that blame
assignment adds value. Similarly, researchers and language designers have
implicitly answered this question one way or another without evidence for the
blame-strategy dimension. The experiment presented in this paper provides some
justification for the programmers' leanings and helps language creators revisit
their decisions. Of course, the design choice remains a trade-off, and the
the experiment contribute to only one of its many facets.

The paper does {\em not\/} address a problem in the gradually typed world that
was pointed out early on by practical researchers~\cite{incorrect-ts,
sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
attention~\cite{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
themselves.  Developers use gradual typing to move an untyped code base into the
typed realm, and to this end, they need typed APIs for the vast repositories of
already-existing libraries. Instead of converting the libraries themselves,
language implementors merely create typed facade modules that import untyped
functions and export them with type annotations, like {\tt typed-pack-lib} in
figure~\ref{fig:example}.  With those facades, the compiler can easily
type-check typed modules. The problem is that these retroactive additions of
types to a library may result from a misunderstanding of the code. \emph{Any
retroactively ascribed type may thus be a mistake itself.}

The cited evidence suggests that this scenario is quite common and largely
unadressed.  A future evaluation must develop mutators that produce incorrect
type annotations without breaking the code itself. Some preliminary work
suggests that such type mutators are even more difficult to develop than the
type-level code mutators presented here. Based on the gradual typing literature,
the Natural semantics should usually discover such mistakes. In contrast, the
Transient and Erasure semantics cannot help with such mistakes at all; indeed,
we expect that these latter two raise misleading exceptions or produce
plainly-wrong answers.  Of course, only additional experimental work can confirm
or reject these conjectures.

### 2021-03-04 writing/discussion.tex
issues with sections 9, 10, and 11, plus triggers for global searches

%% -----------------------------------------------------------------------------
%% \begin{itemize}
%% \item Why does the data look like this?
%%   - why tr fails
%%     - bad runtime errors
%%     - TR bug
%%   - why transient fails
%%     - bad runtime errors
%%     - timeouts/ooms: give context wrt fully untyped/typed versions
%%     - empty blame: trim Ben's prose

%%   \item Any kind of dynamic checking has a large positive impact, whether blame or stacktraces are used.
%%     \begin{itemize}
%%     \item Both in terms of usefulness and programmer effort
%%     \item Erasure extremely often fails to provide any helpful information.
%%     \item That said, there are no long trails for Erasure.
%%     \end{itemize}

%%   \item Blame helps significantly over just using stacktraces for both Natural and Transient.
%%     \begin{itemize}
%%     \item Natural's blame improves over exceptions by nearly 20\%
%%     \item Transient's blame improves over exceptions half as much
%%     \end{itemize}

%%   \item Natural clearly offers the most utility, but Transient is not too far behind.
%%     \begin{itemize}
%%     \item Natural clearly improves over all other modes
%%     \item Transient improves significantly over all other modes except Natural.
%%     \item Keep in mind that our selection of scenarios is biased toward Erasure & Transient.
%%     \end{itemize}

%%   \item Discussion of how confidence / margin of error should be interpreted wrt these results?

%%    \item Threats to validity:
%%         \begin{itemize}
%%           \item Racket stack traces
%%           \item Mutant sampling
%%           \item Blame trails do not necessarily capture all aspects of error reporting
%%           \item Transient blame adaptations
%%           \item More?

%%    \end{itemize}
%% \end{itemize}

The analysis in the previous section suggests a number of interesting
high-level conclusions about blame in the gradual typing setting.  First,
run time type checks have a large positive impact. This is the case
regardless of whether these checks issue blame or throw plain
exceptions.  Second, in many cases type checks that issue blame are more
helpful than those that do not. However, the results indicate that blame
is not critical in a majority of cases, and therefore it is worth further
investigation whether disabling blame tracking for the sake of performance
may be   an acceptable trade-off. Third, from the three different blame
strategies, the Natural approach of Typed Racket fares better than the
Transient approach of Reticulated Python, but only by a small margin.
Since Natural offers complete and sound path-based blame while Transient
offers incomplete but sound heap-based blame~\cite{gfd-oopsla-2019}, the
results call for a deeper understanding of the relative strengths of the
two models for blame.  Fourth, given that Transient's sound but shallow
run time type checks do not seem to hamper debugging, a version of Typed
Racket that disables some of its wrappers may offer an answer to the
well-known performance issues of Natural~\cite{gtnffvf-jfp-2019}.  Fifth,
just as for Typed Racket, the experiment points to possible improvements
for Reticulated Python. In particular, the experimental data indicates
that the usefulness of transient blame does not depend on whether  the
rational programmer follows first or the last element from a blame set,
and that using the last element seems to produce shorter trails.  Hence,
given the performance issues of Transient discussed further on, a possible
optimization is to limit the size of blame sets to only contain the latest
entries using timestamps.

There are a number of threats to the validity of these conclusions: (i) the
representativeness of benchmarks; (ii) the relation between mutations and
real programming mistakes; (iii) the definition of interesting debugging
scenarios; and (iv) the sampling strategy. The previous sections
have described the various ways we attempt to mitigate these threats;
the remainder of this section discusses three additional threats.

@@

Like {\em homo economicus\/}, which decouples the actual behavior of a
participant in an economy for the sake of mathematical modeling, the model of a
rational programmer decouples the actual debugging behavior of a software
developer for the sake of a systematic, large-scale analysis. This decoupling
comes with advantages and disadvantages. In the economic realm, mathematical models
have provided some predictive insights into the market's behavior; but as
behavioral economics has shown more recently, the mathematical abstraction of a
rational actor makes predictions also quite unreliable in some situations.
\footnote{It has also misled economists to focus on just the mathematics, though
this problem is not relevant here.}  Just like an ordinary consumer or producer,
an actual software developer is unlikely to stick to the exact strategy proposed
here. When this happens, the predicted benefits of blame assignment may not
materialize. Indeed, the authors' personal experience suggests such deviations, and
it also suggests that deviating is a mistake. To make a true judgment of the
usefulness of the rational-programmer idea, the community will need much more
experience with this form of evaluation and relating the evaluation to the
behavior of working programmers.

Relatedly, the experimental setup hides the type choice a rational programmer must make. When the
run-time checks signal an impedance mismatch in the real world, a programmer
does not have a typed module ready to swap in. Instead, the programmer must
come up with the next set of types---and this means the programmer must make
choices. It is normally possible to assign consistent types to variables in a
module in different ways. The creation and curation of the benchmarks over
many years has driven home this lesson but, fortunately, it has also shown that
the types are in most cases reasonably canonical.  The authors therefore conjecture that
a rational programmer would in most cases come up with an equivalent type
assignment in order to extend a trail as the experimental method describes.

@@

The results in section~\ref{sec:results} show that Transient produces empty
blame sequences for 967 scenarios. In theory, empty set must not occur. An
empty blame sequence means a lack of boundary crossings for the witness value.
By implication, a typed module is blaming itself---something that can
happen only if the type checker (or system) is unsound.

An investigation of these empty blame cases reveals that Transient is not
unsound but that Vitousek et al.'s Transient blame assignment loses track
of the proper boundaries due to an incomplete population of the blame map.
Despite these losses, the number of failures is a small fraction of the
overall number of debugging scenarios. It is unlikely that any
improvements to Vitousek et al.'s strategy would change the overall
conclusion about Transient blame.

{\bf Note} The scenarios where Transient returns no blame 
suggest ways to improve its blame map:
\begin{itemize}

\item Entries in a blame map should point to {\em several\/} parent entries.
 For example, if \texttt{f} receives bad input in the context of {\tt (filter f
 xs)}, the entry should point to both \texttt{filter} and \texttt{xs}.

\item The construction of entries must be guided by type-like specifications for
 primitives instead of special cases for built-in functions. Aliasing such
 function changes the blame map.

\end{itemize}
This small example is indicative of how rational-programmer experiments may
provide actionable feedback for language design .

@@

At first glance, these measurements seem to contradict the results
of~\citet{vss-popl-2017}. They report an average slowdown of 2.5x and a
worst-case of 5.4x on fully-typed benchmarks in Reticulated Python.
Unfortunately, the high cost of Transient Racket seems closer to the truth.
There are at least four broad issues that skew the earlier results:
\begin{enumerate}

@@

\item On large programs, Reticulated suffers from high overhead.  For example, a
Reticulated version of the simplest benchmark runs in ~40 seconds without blame,
and with blame times out after 10 minutes. Typed Racket runs the same benchmark
in a small fraction of the time.
\end{enumerate}
More work on Transient blame is needed to make an informed decision.

>>>>

%% -----------------------------------------------------------------------------

The results of the experiment suggests a number of high-level conclusions about
blame strategies in the gradual typed world.  First, run-time type checks have a
large positive impact, regardless of whether these checks assign blame or throw
plain exceptions.  Second, error messages with blame assignments are more
helpful than those without. The results also indicate, though, that blame is not
critical in a majority of cases, and therefore they suggest investigating
whether blame tracking is worth the performance cost.  Third, the Natural
approach fares better than the Transient approach, but only by a small margin.
Since Natural offers complete and sound path-based blame while Transient offers
incomplete but sound heap-based blame~\cite{gfd-oopsla-2019}, the results call
for a study concerning the relative strengths of the two models of blame.
Fourth, given that Transient's sound but shallow run-time type checks do not
seem to hamper debugging, a version of the Natural semantics of Typed Racket
that disables some of its wrappers may offer an answer to the well-known
performance issues of Natural~\cite{gtnffvf-jfp-2019}.  Fifth, given that both
modes of the Transient rational programmer are equally successful and that
Transient-last produces shorter trails, Reticulated Python could limit the size
of blame sequences to the last few entries to reduce its serious performance
problems (see below).

The validity of these conclusions is threatened in several ways: (i) the
representativeness of benchmarks; (ii) the relation between mutations and real
programming mistakes; (iii) the definition of interesting debugging scenarios;
and (iv) the sampling strategy. While these threats have been mitigated by the
experimental setup, the remainder of this section discusses three additional
ones.

@@

Like {\em homo economicus\/}, which decouples the actual behavior of a
participant in an economy for the sake of mathematical modeling, the model of a
rational programmer decouples the actual debugging behavior of a software
developer for the sake of a systematic, large-scale analysis. This decoupling
comes with advantages and disadvantages. In the economic realm, mathematical
models have provided some predictive insights into the market's behavior; but as
behavioral economics has shown more recently, the mathematical abstraction of a
rational actor makes predictions also quite unreliable in some situations.
\footnote{It has misled economists to focus on just mathematics, though
this problem is not relevant here---tongue firmly in cheek.}  Just like an
ordinary consumer or producer, an actual software developer is unlikely to stick
to the exact strategy proposed here. When this happens, the predicted benefits
of blame assignment may not materialize. Indeed, the authors' personal
experience suggests such deviations, and it also suggests that deviating is a
mistake. To make a true judgment of the usefulness of the rational-programmer
idea, the community will need much more experience with this form of evaluation
and relating the evaluation to the behavior of working programmers.

Relatedly, the experimental setup hides how a rational programmer ascribes types
to extend a trail. When the run-time checks signal an impedance mismatch in the
real world, a real-world programmer does not have a typed module ready to swap
in. Instead, the programmer must come up with the next set of types, which means
making choices. It is usually possible to consistently assign types to variables
in a module in different ways. The maintenance of the benchmarks over many years
has driven home this lesson but, fortunately, it has also shown that the types
are in somewhat canonical.  The authors therefore conjecture that different
real-world programmers would often come up with equivalent type assignments
during debugging sessions.

@@

The execution of the experiment reveals that Transient produces empty blame
sequences for 967 scenarios. In theory, empty set must not occur. An empty blame
sequence means a lack of boundary crossings for the witness value.  By
implication, a typed module is blaming itself---something that can happen only
if the type checker (or system) is unsound.

An investigation of these empty blame cases reveals that Transient is not
unsound but that Vitousek et al.'s Transient blame assignment loses track
of the proper boundaries due to an incomplete initialization of the blame map.
Despite these losses, the number of failures is a small fraction of the
overall number of debugging scenarios. It is unlikely that any
improvements to Vitousek et al.'s strategy would change the overall
conclusion about Transient blame.

{\bf Note} The scenarios where Transient returns no blame suggest two ways to
improve its blame map.  First, entries in a blame map should point to {\em
several\/} parent entries.  For example, if \texttt{f} receives bad input in the
context of {\tt (filter f xs)}, the entry should point to both \texttt{filter}
and \texttt{xs}. Second, the construction of entries must be guided by type-like
specifications for primitives instead of special cases for built-in
functions. Aliasing such function changes the blame map.

@@

At first glance, these measurements seem to contradict the results
of~\citet{vss-popl-2017}. They report an average slowdown of 2.5x and a
worst-case of 5.4x on fully-typed benchmarks in Reticulated Python.
Unfortunately, the high cost of Shallow Racket seems closer to the truth.
There are at least four broad issues that skew the earlier results:
\begin{enumerate}

@@

\item On non-trivial programs, Reticulated Python suffers from the same high
overhead as Shallow Racket.  It runs the simplest benchmark from the suite 
in approximately 40 seconds without blame and times out after 10 minutes when
the blame map is enabled.

\end{enumerate}
More work on Transient blame is needed to make an informed decision.

### 2021-03-04 writing/mutate-mutators.tex
issues with sections 9, 10, and 11, plus triggers for global searches

Most of the mutators are self-explanatory.  The first six apply to all
gradually typed languages; the next six to those that include classes and
objects. The last four target distinguishing features of Typed Racket,
specifically its sophisticated type system. For example, one mutation produced by {\tt arithmetic}
replaces a {\tt +} with a {\tt -} in an attempt to change the type of the
arithmetic expression; {\tt +}'s result is a {\tt Positive-Integer} when all
arguments are positive integers, while {\tt -} yields {\tt
Integer}~\cite{stathff-padl-12}. Similarly, the other three aim to confound
Typed Racket's occurrence type system.

>>>>

Most of the mutators are self-explanatory.  The first six apply to all
gradually typed languages; the next six to those that include classes and
objects. The last four target distinguishing features of Typed
Racket's type system,
specifically its sophisticated type system. For example, one mutation produced by {\tt arithmetic}
replaces a {\tt +} with a {\tt -} in an attempt to change the type of the
arithmetic expression; {\tt +}'s result is a {\tt Positive-Integer} when all
arguments are positive integers, while {\tt -} yields {\tt
Integer}~\cite{stathff-padl-12}. Similarly, the other three aim to confound
the occurrence type system.

### 2021-03-04 writing/rational-modes.tex
issues with sections 9, 10, and 11, plus triggers for global searches

While a successful Natural blame trail indicates that it 
pays off to heed blame assignments during debugging the trail's root, it does not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run time type check, dubbed the
location of the exception, might be as useful as typing the blamed one.

>>>>

While a successful Natural blame trail indicates that it 
pays off to heed blame assignments during debugging the trail's root, it does not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run-time type check, dubbed the
location of the exception, might be as useful as typing the blamed one.

### 2021-03-04 writing/related.tex
issues with sections 9, 10, and 11, plus triggers for global searches

The literature on gradual typing presents many semantics beyond the three used
here and only two additional blame strategies.  Pyret
(\shorturl{https://www.}{pyret.org}) assigns fixed-size data types the Natural
semantics and functions a Transient semantics. The Forgetful~\cite{cl-icfp-2017}
and the Amnesic~\cite{gfd-oopsla-2019} semantics are similar to Transient but
use wrappers instead of in-lined checks.  Nom~\cite{mt-oopsla-2017} and other
\emph{concrete\/} semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015,
rzv-ecoop-2015, rat-oopsla-2017} assume that every value comes with a type tag
and use tag checks to supervise the interactions between typed and untyped code.
Monotonic references~\cite{svctg-esop-2015} and the
semantics~\cite{tlt-popl-2019, etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17}
derived with the Abstracting Gradual Typing technique~\cite{gct-popl-2016} are
variants of Natural.  Among these representative semantics, only Amnesic and Nom
present innovative blame strategies.  The experiment excludes these strategies
because they seem to be merely theoretical constructions and/or impose severe
restrictions on programmers.

>>>>

The literature on gradual typing presents many semantics beyond the three used
here and two additional blame strategies.  Pyret (pyret.org) assigns fixed-size
data types the Natural semantics and functions a Transient semantics. The
Forgetful~\cite{cl-icfp-2017} and the Amnesic~\cite{gfd-oopsla-2019} semantics
are similar to Transient but use wrappers instead of in-lined checks.
Nom~\cite{mt-oopsla-2017} and other \emph{concrete\/}
semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
rat-oopsla-2017} assume that every value comes with a type tag and use tag
checks to supervise the interactions between typed and untyped code.  Monotonic
references~\cite{svctg-esop-2015} and the semantics~\cite{tlt-popl-2019,
etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
Gradual Typing technique~\cite{gct-popl-2016} are variants of Natural.  Among
these representative semantics, only Amnesic and Nom present innovative blame
strategies.  The experiment excludes these strategies because they seem to be
merely theoretical constructions and/or impose severe restrictions on
programmers.

### 2021-03-04 writing/results.tex
issues with sections 9, 10, and 11, plus triggers for global searches

While most blame trail failures follow the above pattern, a few do
not.  Breaking down the failure reasons for Natural blame (1748 in
total) reveals an additional cause. For a small set of debugging scenarios
(40), Natural produces a run time type error blaming a non-buggy
already-typed module. All these cases are due to known open issues with Typed
Racket and class contracts.

@@

Second, the exceptions of the Erasure semantics either help the rational
programmer immediately or the rational programmer fails to debug a scenario
altogether (in red).  This is expected; adding type annotations to an Erasure
program does not change its behavior, so if the type checker does not reject the
program with the new annotations, the rational programmer is stuck.  In other
words, an exception from the runtime has to point to the buggy module with the
first try. Otherwise the rational programmer types an irrelevant module, runs
the program again, and the same exception points again to the already typed
module.

>>>>

While most blame trail failures follow the above pattern, a few do
not.  Breaking down the failure reasons for Natural blame (1748 in
total) reveals an additional cause. For a small set of debugging scenarios
(40), Natural produces a run-time type error blaming a non-buggy
already-typed module. All these cases are due to known open issues with Typed
Racket and class contracts.

@@

Second, the exceptions of the Erasure semantics either help the rational
programmer immediately or the rational programmer fails to debug a scenario
altogether (in red).  This is expected; adding type annotations to an Erasure
program does not change its behavior, so if the type checker does not reject the
program with the new annotations, the rational programmer is stuck.  In other
words, an exception from the run-time system has to point to the buggy module with the
first try. Otherwise the rational programmer types an irrelevant module, runs
the program again, and the same exception points again to the already typed
module.
